Follow-up Samples

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add pdf with completed results

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  html_document: default
geometry: margin=1.0in
bibliography: reference.bib
---

@@

\vspace{25mm}

\begin{center}
Running Title: Human Microbiome and Colorectal Cancer

@@

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between stool bacterial microbiome and CRC.   The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.   
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.  Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and 'r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`  
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(pvalue_adn_crc_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(pvalue_adn_crc_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4]. Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

@@

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  Our middle training model, based on AUC, from 100 80/20 (train/test) splits was similar to the full training data model.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using the full training data model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.  Although there was a decrease in overall probability of an adenoma or carcinoma only `r followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S3].  Many of the OTUs identified classified to normal flora bacterium [Table S3].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.  We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

>>>>

---
title: ""
bibliography: reference.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  html_document: default
csl: microbiome.csl
fontsize: 12pt
geometry: margin=1.0in
---

@@

```
\begin{center}  
\vspace{25mm}

Running Title: Human Microbiome and Colorectal Cancer

@@

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between stool bacterial microbiome and CRC.   The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.   
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.  Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

***Differences in Adenoma and Carcinoma of Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with Porphyromonas asaccharolytica (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), Fusobacterium nucleatum (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), Parvimonas micra (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and Peptostreptococcus stomatis (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for Peptostreptococcus stomatis (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`) and Porphyromonas asaccharolytica (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no signficant differences in any of these OTUs in the adenoma samples [Table S4]. Second, when these OTUs were present, there was a clear magnitude difference in these specific OTUs based on whether they were from adenoma or carcinoma individuals [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

@@

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  Our middle training model, based on AUC, from 100 80/20 (train/test) splits was similar to the full training data model.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using the full training data model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample was a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was `r #format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r #format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.  Although there was a decrease in overall probability of an adenoma or carcinoma only `r #followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r #sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S3].  Many of the OTUs identified classified to normal flora bacterium [Table S3].  Only a single OTU though was significant after multiple comparison correction and the lowest taxonomic identification of `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r #filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  Although we were interested in what we could use to classify those with either adenoma or carcinoma versus normal.  We found that the traditional bacteria associated with CRC were higher in magnitude in the carcinoma group and there were significant differences in some of these OTUs between the initial and follow up samples [Figure 5 and Tabl S5].  This research provides evidence that it is possible to use bacterial microbiome data to create a highily sensitive model, that is reactive to therapy, for detection of adenoma or carcinoma.  It accomplishes this by using a unique sample set in which before and after surgery stool samples are available for assessment.  By using these types of samples we are not only able to show sensitivity of lesion prediction but also able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.
